---
title: "HW6-W43-functions-gapminder"
author: "Mina Almasi"
date: '2022-10-30'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setting up
```{r}
library(gapminder)
library(tidyverse)
```

```{r}
data <- gapminder
```

## Exercise 1: Writing a Defensive Function
Define a defensive function that calculates the Gross Domestic Product of a nation from the data available in the gapminder dataset. You can use the population and GDPpercapita columns for it. Using that function, calculate the GDP of Denmark in the following years: 1967, 1977, 1987, 1997, 2007, and 2017.
```{r}
gdp_calculator <- function(data, year, country) {
  
  # selecting relevant row in data 
  data <- data[which(data$country == country & data$year == year), ]
  
  #calculating gdp 
  gdp <- round(data$pop * data$gdpPercap)
  
  # printing GDP if row exists (if there is a row i.e., a country and a year, then it will print the GDP)
  ifelse(nrow(data) == 1, paste(country, "has a GDP of", gdp, "in international dollars"), "MISSING VALUE: COUNTRY OR YEAR")
}
```

With the function that is defined, we calculate try to calculate the years 1967, 1977, 1987, 1997, 2007, and 2017.
```{r}
years <- c(1967, 1977, 1987, 1997, 2007, 2017)

for (i in 1:length(years)){
  print(gdp_calculator(gapminder, year = years[i], country = "Denmark"))
}
```


## Exercise 2: Script 
Write a script that loops over each country in the gapminder dataset, tests whether the country starts with a ‘B’ , and prints out whether the life expectancy is smaller than 50, between 50 and 70, or greater than 70. (Hint: remember the grepl function, and review the Control Flow tutorial)

Assuming that we are meant to test only (and not filter upon) countries starting with B !! The loop will thus print all countries but denote whether it starts with B or not. 
```{r}
# defining B-countries 
B_countries = grep("^B", unique(gapminder$country), value = TRUE)

# list of all unique countries in the data
all_countries = unique(data$country)

#starting loop 
for(country_i in all_countries) {
  temp = data %>% filter(gapminder$country == country_i) #temp data with only the country in the loop (country_i)
  mean_lifeEXP = round(mean(temp$lifeExp), 2) #calculate the mean life expectancy for country_i

  #check whether b is the start letter
  if (country_i %in% B_countries){
    b_start = "the country starts with B"
  }
  else {
    b_start = "the country does NOT start with B"
  }
  
  # check whether mean is less than 50, greater than 50, between 50 and 70
  if (mean_lifeEXP < 50) {
    mean_message = "less than 50"
  }
  if (mean_lifeEXP > 50 & mean_lifeEXP < 70){
    mean_message = "between 50 and 70"
  }
  if (mean_lifeEXP > 70) {
    mean_message = "greater than 70"
  }
  
  output = paste0(country_i, " has a mean life expectancy that is ", mean_message, " and ", b_start)
  
  print(output)
}
```


